---
title: "Titanic"
author: "Tomas Bartos"
date: "13 prosince 2017"
output: 
  html_document:
    fig_height: 4.5
    fig_width: 7
    number_sections: yes
    toc: yes
---

# Introduction

This is my first Kaggle script. 


## Loading libraries:

```{r, message=FALSE, warning = FALSE}
# Load packages
library('readr') # data input
library('VIM') # data input
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('PASWR2') # visualization
library('dplyr') # data manipulation
library('purrr') # data manipulation
library('mice') # imputation
library('randomForest') # classification algorithm
```

## Loading data:

```{r, message=FALSE, warning=FALSE}
# load data
train <- read_csv('train.csv')
test  <- read_csv('test.csv')

# combine training + test data
titanic_original  <- full_join(train, test) 
titanic  <- full_join(train, test)

# factor variables
titanic <- titanic %>% mutate(
  Survived = factor(Survived),
  Pclass = factor(Pclass),
  Embarked = factor(Embarked),
  Sex = factor(Sex)
)

```

Using readr package enable to load data as tibble dataframe, which is  enhanced version of the standard data frame.
Using full_join from dplyr package returns all data from both datasets combining the data from the matched keys

Now we're working with 1309 observations of these variables:

Variable Name | Description
--------------|-------------------------------------------------
Survived      | Survived (1) or died (0)
Pclass        | Passenger's class (1,2,3)
Name          | Passenger's name
Sex           | Passenger's sex - male (0), female (1)
Age           | Passenger's age
SibSp         | Number of siblings/spouses aboard
Parch         | Number of parents/children aboard
Ticket        | Ticket number
Fare          | Fare
Cabin         | Cabin
Embarked      | Port of embarkation (S = Southampton, C = Cherbourg, Q = Queenstown)


## Data overview:
```{r, message=FALSE, warning=FALSE}

glimpse(titanic)
summary(titanic)

```

## Missing data:

Map functions transform their input by applying a function to each element and returning a list or vector the same length as the input.

Using aggr from VIM package plot missing values in each variable and the amount of missing values in certain combinations of variables. 

```{r, message=FALSE, warning=FALSE}

map_int(titanic, ~sum(is.na(.)))

aggr(titanic, col=(c("lightgray", "firebrick")), prop = FALSE, 
     combined = TRUE, numbers = TRUE, sortVars = TRUE, sortCombs = TRUE)



```


# Exploration, Visualisation and Imputation

Data exploration is the first step in data analysis and typically involves summarizing the main characteristics of a dataset.

## Variables

### Gender

```{r, message=FALSE, warning=FALSE}

gender_bar <- ggplot(titanic, aes(x=Sex)) +
  geom_bar(fill=c("firebrick", "dodgerblue4")) +
  ggtitle("Gender counts\nall passengers")
gender_prop <- ggplot(data = titanic[!is.na(titanic$Survived),],  aes(x=Sex,  fill=Survived)) +
  geom_bar(position="fill") +
  scale_fill_manual(values= c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Gender") +
  ylab("Survival Rate") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) +
  annotate("text", x = "male", y = 0.5, label = "Overall\nSurvival\nRatio", colour='white')
multiplot(gender_bar, gender_prop, layout=matrix(c(1,1,2,2,2),nrow = 1, ncol = 5, byrow=TRUE)) 

```

### P Class

```{r, message=FALSE, warning=FALSE}

pclass_bar <- ggplot(titanic, aes(x=Pclass)) +
  geom_bar(fill=c("gold", "darkgray", "chocolate")) +
  ggtitle("Passenger's class counts\nall passengers")
pclass_prop <- ggplot(data = titanic[!is.na(titanic$Survived),],  aes(x=Pclass,  fill=Survived)) +
  geom_bar(position="fill") +
  scale_fill_manual(values= c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Passenger's class") +
  ylab("Survival Rate") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) + 
  annotate("text", x = 3, y = 0.5, label = "Overall\nSurvival\nRatio", colour='white')
multiplot(pclass_bar, pclass_prop, layout=matrix(c(1,1,2,2,2),nrow = 1, ncol = 5, byrow=TRUE)) 

```


### Fare

```{r, message=FALSE, warning=FALSE}
fare_dens <- ggplot(titanic, aes(Fare)) +                       
  geom_density(fill="dodgerblue4", alpha=0.5) +
  geom_vline(aes(xintercept=median(Fare, na.rm=TRUE)), colour='black', lty=2, lwd=1) +
  annotate("text", x = 40, y = 0.04, label = "Median", colour='black') +
  geom_vline(aes(xintercept=mean(Fare, na.rm=TRUE)), colour='firebrick', lty=2, lwd=1) +
  annotate("text", x = 55, y = 0.03, label = "Mean", colour='firebrick') +
  ggtitle("Fare distribution - all known passengers")

fare_hist <- ggplot(data = titanic[!is.na(titanic$Survived),],  aes(x = Fare,  fill = Survived)) +
  geom_histogram(position = "fill", binwidth = 30) +
  scale_fill_manual(values = c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Fare") +
  ylab("Survival Rate") +
  theme(legend.position = "bottom") +
  geom_vline(aes(xintercept = median(Fare, na.rm=TRUE)), colour = 'black', lty = 2, lwd = 1) +
  annotate("text", x = 60, y = 0.1, label = "Median Fare", colour='black') +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) +
  annotate("text", x = 200, y = 0.45, label = "Overall Survival Ratio", colour='white')
multiplot(fare_dens, fare_hist, layout=matrix(c(1,2,2),nrow = 3, ncol = 1, byrow=TRUE)) 
```

Ony 1 Fare values is missing. Let's see this passenger:

```{r, message=FALSE, warning=FALSE}
# Extract the passengers data which contains the missing Fare data
data.frame(filter(titanic, is.na(Fare)))
```

We can assume, that this passenger paid typical fare for third class ticket and for embarkation in Southampton.

```{r, message=FALSE, warning=FALSE}
# Imputation the missing Fare value
titanic$Fare[which(is.na(titanic$Fare))] <- titanic %>% filter(Pclass==3, Embarked=="S") %>% select(Fare) %>% unlist %>% median(na.rm=TRUE)
```

### Embarked

```{r, message=FALSE, warning=FALSE}

embarked_bar <- ggplot(data = titanic[!is.na(titanic$Embarked),], aes(x=Embarked, fill = Embarked)) +
  geom_bar() +
  scale_fill_brewer(labels=c("Cherbourg", "Queenstown", "Southampton")) +
  theme(legend.position = c(0.3, 0.7)) +
  ggtitle("Embarked counts\nall known passengers")
embarked_prop <- ggplot(data = titanic[!is.na(titanic$Survived) & !is.na(titanic$Embarked),],  aes(x=Embarked,  fill=Survived)) +
  geom_bar(position="fill") +
  scale_fill_manual(values= c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Embarked data") +
  ylab("Survival Rate") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) + 
  annotate("text", x = 3, y = 0.5, label = "Overall\nSurvival\nRatio", colour='white')
multiplot(embarked_bar, embarked_prop, layout=matrix(c(1,1,2,2,2),nrow = 1, ncol = 5, byrow=TRUE)) 

```

We have noticed previously that there are two missing values for the Embarked variable.

```{r, message=FALSE, warning=FALSE}
# Extract the passengers data which contains the missing Embarked data
data.frame(filter(titanic, is.na(Embarked)))
```

We can see, that Southampton port is the most frequent port of embarkation but we have to take into account also fare (80) and passengers class (1) of these female passengers. 

```{r, message=FALSE, warning=FALSE}
titanic %>% 
  filter(Pclass == 1 & !is.na(Embarked)) %>% 
  ggplot(aes(x = Embarked, y = Fare, fill = Embarked)) + 
    geom_boxplot() +
    scale_fill_brewer() +
    geom_hline(aes(yintercept = 80), colour = "red", lty = 2, size = 1) +
    annotate("text", x = 2, y = 35, label = "Fare paid by\n2 female passengers\nwith missing embarked location", colour = "red") +
    ggtitle("Fare distribution of first class passengers") 


```

From boxplot is obvious that Fare values about 80 is typical for Cherbourg port class 1 passengers, so We can assume, that these passengers embarked Cherbourg port.

```{r, message=FALSE, warning=FALSE}
# Imputation the missing Embarked value
titanic$Embarked[which(is.na(titanic$Embarked))] <- "C"
```

### Passenger Title - engineered feature

The passenger title (“Mrs”, “Master”, “Mr”, etc.) is part of Passenger’s name variable . Certain titles can be helpful for later missing age data imputation.

```{r, message=FALSE, warning=FALSE}
# Grabing passenger title from passenger name variable
titanic$Title <- gsub("^.*, (.*?)\\..*$", "\\1", titanic$Name)

# Title counts by sex
table(titanic$Sex, titanic$Title)

# Reassign some categories 
titanic$Title[titanic$Title == 'Mlle' | titanic$Title == 'Ms'] <- 'Miss' 
titanic$Title[titanic$Title == 'Mme']  <- 'Mrs' 
titanic$Title[titanic$Title %in% c('Dona', 'Dr', 'Lady', 'the Countess','Capt', 'Col', 'Don', 'Jonkheer', 'Major', 'Rev', 'Sir')]  <- 'Other'

# Let's see results
titanic$Title <- factor(titanic$Title, levels = c("Miss", "Mrs", "Master", "Mr", "Other"))
table(titanic$Sex, titanic$Title)

title_bar <- ggplot(data = titanic, aes(x=Title, fill=Title)) +
  geom_bar() +
  coord_flip() +
  scale_fill_brewer(palette = "Spectral") +
  ggtitle("Title counts\nall passengers")
title_prop <- ggplot(data = titanic[!is.na(titanic$Survived),],  aes(x=Title,  fill=Survived)) +
  geom_bar(position="fill") +
  scale_fill_manual(values= c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Title data") +
  ylab("Survival Rate") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) + 
  annotate("text", x = 4, y = 0.48, label = "Overall\nSurvival\nRatio", colour='white')
multiplot(title_bar, title_prop) 
```

### Family size - engineered feature

Using SigSp and Parch datacan be used to calculation Family Size variable as follows:

```{r, message=FALSE, warning=FALSE}
titanic$FamilySize <- 1 + titanic$SibSp + titanic$Parch
table(titanic$FamilySize)


family_bar <- ggplot(data = titanic, aes(x=FamilySize, fill=factor(FamilySize))) +
  geom_bar() +
  scale_x_continuous(breaks = c(1:8, 11)) +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Spectral") +
  xlab("Family size (members)") +
  ggtitle("Family Size counts all passengers")
family_prop <- ggplot(data = titanic[!is.na(titanic$Survived),],  aes(x=FamilySize,  fill=Survived)) +
  geom_bar(position="fill") +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = c(1:8, 11)) +
  scale_fill_manual(values= c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Family Size") +
  ylab("Survival Rate") +
  xlab("Family size (members)") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) + 
  annotate("text", x = 6.5, y = 0.6, label = "Overall Survival Ratio", colour='white')
multiplot(family_bar, family_prop) 
```

It is obvious that some family size habe similar survival rates, so we can simplify these data by engineering new factor variable.

```{r, message=FALSE, warning=FALSE}
# Let's simplify data via category
titanic$Fsize <- titanic$FamilySize %>% map_chr(function(x) ifelse(x == 1, "Single", ifelse(x> 7, "Extreme", ifelse(x> 4, "Large", "Small")))) %>% factor(levels = c("Single", "Small", "Large", "Extreme"), ordered = TRUE)
table(titanic$Fsize)

fsize_bar <- ggplot(data = titanic, aes(x=Fsize, fill=Fsize)) +
  geom_bar() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Spectral") +
  xlab("Family size") +
  ggtitle("Family Size counts all passengers")
fsize_prop <- ggplot(data = titanic[!is.na(titanic$Survived),],  aes(x=Fsize,  fill=Survived)) +
  geom_bar(position="fill") +
  theme(legend.position = "bottom") +
  scale_fill_manual(values= c("grey25", "darkolivegreen3")) +
  ggtitle("Survival Rate by Family Size") +
  ylab("Survival Rate") +
  xlab("Family size") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) + 
  annotate("text", x = 3, y = 0.6, label = "Overall Survival Ratio", colour='white')
multiplot(fsize_bar, fsize_prop) 
```


### Age

```{r, message=FALSE, warning=FALSE}
age_dens <- ggplot(titanic[!is.na(titanic$Age),], aes(Age)) +                
  geom_density(fill="dodgerblue4", alpha=0.5) +
  scale_x_continuous(breaks = seq(5,80,5)) +
  ggtitle("Age distribution - all known passengers")

age_hist <- ggplot(data = titanic[!is.na(titanic$Survived) & !is.na(titanic$Age),],  aes(x = Age,  fill = Survived)) +
  geom_histogram(position = "fill", binwidth = 5) +
  scale_fill_manual(values = c("grey25", "darkolivegreen3")) +
  scale_x_continuous(breaks = seq(5,80,5)) +
  ggtitle("Survival Rate by Age - all known passengers") +
  ylab("Survival Rate") +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) +
  annotate("text", x = 40, y = 0.55, label = "Overall Survival Ratio", colour='white')
multiplot(age_dens, age_hist, layout=matrix(c(1,2,2),nrow = 3, ncol = 1, byrow=TRUE)) 
```
 

Now we can have a look if this survival rate versis age data is biased, for example known age data is more abundant in survived passengers.

```{r, message=FALSE, warning=FALSE} 
# Engineered new variable - known/unknown Age data
titanic$age_known <- factor(ifelse(is.na(titanic$Age), "unknown", "known"))

table(titanic$age_known)

ggplot(titanic[!is.na(titanic$Survived),], aes(x = age_known, fill = Survived)) +                
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("grey25", "darkolivegreen3")) +
  geom_hline(yintercept = (sum(train$Survived)/nrow(train)), col = "white", lty = 2, lwd = 1) +
  annotate("text", x = 1, y = 0.55, label = "Overall Survival Ratio", colour='white')
```
 
It is aparent that survival ratio for passengers with known Age data is close to overall survival rate. That is primarily due to high proportion of known Age data. We can conclude that presented Age distribution histogram is relatively representative, histogram with all Age data will be very similar, survival rates for different age intervals will be only insignificantly lower due to higher propoertion of unsurvived passengers.
 
   
We have noticed that there are many missing values for the Age variable.          
```{r, message=FALSE, warning=FALSE} 
# Imputation missing data
mice_age <- mice.impute.rf(titanic$Age, !is.na(titanic$Age), titanic[, c("Title","Pclass", "Sex", "SibSp", "Parch", "Fare", "Embarked")]) 

titanic$Age[is.na(titanic$Age)] <- mice_age


ggplot(titanic, aes(x = Age, fill = age_known)) +                
  geom_histogram(position="identity", alpha=0.6) +
  scale_x_continuous(breaks = seq(5,80,5)) +
  labs(fill ="Age") +
  theme(legend.position = c(0.85, 0.8)) +
  ggtitle("Age distribution - (un)known passengers")

```

4 Prediction
At last we’re ready to predict who survives among passengers of the Titanic based on variables that we carefully curated and treated for missing values. For this, we will rely on the randomForest classification algorithm; we spent all that time on imputation, after all.

4.1 Split into training & test sets
Our first step is to split the data back into the original test and training sets.

# Split the data back into a train set and a test set
train_new <- titanic[1:891,]
test_new <- titanic[892:1309,]

4.2 Building the model
We then build our model using randomForest on the training set.

# Set a random seed
set.seed(754)

# Build the model (note: not all possible variables are used)
rf_model <- randomForest(factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize, data = train_new)

# Show model error
plot(rf_model, ylim=c(0,0.36))
legend('topright', colnames(rf_model$err.rate), col=1:3, fill=1:3)

Variable importance
Let’s look at relative variable importance by plotting the mean decrease in Gini calculated across all trees.

4.4 Prediction!
We’re ready for the final step — making our prediction! When we finish here, we could iterate through the preceding steps making tweaks as we go or fit the data using different models or use different combinations of variables to achieve better predictions. But this is a good starting (and stopping) point for me now.

# Predict using the test set
prediction <- predict(rf_model, test_new)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
solution <- data.frame(PassengerID = test$PassengerId, Survived = prediction)

# Write the solution to file
write.csv(solution, file = 'TB_solution.csv', row.names = FALSE)

